---
title: "vitD normalization by clinic+year and SEASON"
author: "David Hughes"
date: "2022-08-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(mgcv)
library(patchwork)

## Load Parameter Data
source("parameters/pfile.txt")
## Load data
f = paste0(datadir, working_data_file_newcolnames)
mydata = read.table(f, header = TRUE, sep = "\t", as.is = TRUE)
```


# Data Filtering

## Remove those with no genetic data

```{r}
w = which( is.na( mydata$pgs ) )
mydata = mydata[-w, ]
dim(mydata)
```
## Keep those with VitD data in at least ONE clinic

```{r}
## Turn all the negative values into NAs
temp = mydata[, c("VitDt_FOM1","VitDt_FOM2","VitDt_FOM3","VitDt_FOM4")]
temp[temp<0] = NA
mydata[, c("VitDt_FOM1","VitDt_FOM2","VitDt_FOM3","VitDt_FOM4")] = temp

## For each individual how many clinics do they have no data for?
no_data_count = apply(mydata[, c("VitDt_FOM1","VitDt_FOM2",
                 "VitDt_FOM3","VitDt_FOM4")], 1, function(x){
  sum( is.na(x) , na.rm = TRUE )
})

## Remove those sample where no_data_count == 4
## indicating that there is no data in any one of the 4 clinics
w = which(no_data_count == 4)
if(length(w)>0){ mydata = mydata[-w,] }
dim(mydata)
```

# BUILD A SEASON Factor Variable

## remove negative values for month

```{r}
mydata$month_attended_FOM1[mydata$month_attended_FOM1 < 0] = NA
mydata$month_attended_FOM2[mydata$month_attended_FOM2 < 0] = NA
mydata$month_attended_FOM3[mydata$month_attended_FOM3 < 0] = NA
mydata$month_attended_FOM4[mydata$month_attended_FOM4 < 0] = NA

mydata$year_attended_FOM1[mydata$year_attended_FOM1 < 0] = NA
mydata$year_attended_FOM2[mydata$year_attended_FOM2 < 0] = NA
mydata$year_attended_FOM3[mydata$year_attended_FOM3 < 0] = NA
mydata$year_attended_FOM4[mydata$year_attended_FOM4 < 0] = NA

```


## FOM1

```{r}
## Build a season variable
r = which( is.na(mydata$month_attended_FOM1) )

mydata$FOM1_season = "autumn"

w = which(mydata$month_attended_FOM1>=1 & mydata$month_attended_FOM1<=3)
mydata$FOM1_season[w] = "winter"

w = which(mydata$month_attended_FOM1>=4 & mydata$month_attended_FOM1<=6)
mydata$FOM1_season[w] = "spring"

w = which(mydata$month_attended_FOM1>=7 & mydata$month_attended_FOM1<=10)
mydata$FOM1_season[w] = "summer"

if(length(r)>0){mydata$FOM1_season[r] = NA}

table(mydata$FOM1_season)
```

## FOM2

```{r}
## Build a season variable
r = which( is.na(mydata$month_attended_FOM2) )

mydata$FOM2_season = "autumn"

w = which(mydata$month_attended_FOM2>=1 & mydata$month_attended_FOM2<=3)
mydata$FOM2_season[w] = "winter"

w = which(mydata$month_attended_FOM2>=4 & mydata$month_attended_FOM2<=6)
mydata$FOM2_season[w] = "spring"

w = which(mydata$month_attended_FOM2>=7 & mydata$month_attended_FOM2<=10)
mydata$FOM2_season[w] = "summer"

if(length(r)>0){mydata$FOM2_season[r] = NA}

table(mydata$FOM2_season)
```

## FOM3

```{r}
## Build a season variable
r = which( is.na(mydata$month_attended_FOM3) )

mydata$FOM3_season = "autumn"

w = which(mydata$month_attended_FOM3>=1 & mydata$month_attended_FOM3<=3)
mydata$FOM3_season[w] = "winter"

w = which(mydata$month_attended_FOM3>=4 & mydata$month_attended_FOM3<=6)
mydata$FOM3_season[w] = "spring"

w = which(mydata$month_attended_FOM3>=7 & mydata$month_attended_FOM3<=10)
mydata$FOM3_season[w] = "summer"

if(length(r)>0){mydata$FOM3_season[r] = NA}

table(mydata$FOM3_season)
```


## FOM4

```{r}
## Build a season variable
r = which( is.na(mydata$month_attended_FOM4) )

mydata$FOM4_season = "autumn"

w = which(mydata$month_attended_FOM4>=1 & mydata$month_attended_FOM4<=3)
mydata$FOM4_season[w] = "winter"

w = which(mydata$month_attended_FOM4>=4 & mydata$month_attended_FOM4<=6)
mydata$FOM4_season[w] = "spring"

w = which(mydata$month_attended_FOM4>=7 & mydata$month_attended_FOM4<=10)
mydata$FOM4_season[w] = "summer"

if(length(r)>0){mydata$FOM4_season[r] = NA}

table(mydata$FOM4_season)
```
## order the SEASON factor

```{r}
mydata$FOM1_season = factor(mydata$FOM1_season, levels = c("winter","spring","summer","autumn"))
mydata$FOM2_season = factor(mydata$FOM2_season, levels = c("winter","spring","summer","autumn"))
mydata$FOM3_season = factor(mydata$FOM3_season, levels = c("winter","spring","summer","autumn"))
mydata$FOM4_season = factor(mydata$FOM4_season, levels = c("winter","spring","summer","autumn"))
```



## Long Format Data

```{r}
fom1 = mydata[, c("aln", "month_attended_FOM1", "year_attended_FOM1", "FOM1_season", "VitDt_FOM1" )  ]
fom1$clinic = "FOM1"
colnames(fom1) = c("aln","month","year","season","vitD","clinic")

fom2 = mydata[, c("aln", "month_attended_FOM2", "year_attended_FOM2", "FOM2_season", "VitDt_FOM2" )  ]
fom2$clinic = "FOM2"
colnames(fom2) = c("aln","month","year","season","vitD","clinic")

fom3 = mydata[, c("aln", "month_attended_FOM3", "year_attended_FOM3", "FOM3_season", "VitDt_FOM3" )  ]
fom3$clinic = "FOM3"
colnames(fom3) = c("aln","month","year","season","vitD","clinic")

fom4 = mydata[, c("aln", "month_attended_FOM4", "year_attended_FOM4", "FOM4_season", "VitDt_FOM4" )  ]
fom4$clinic = "FOM4"
colnames(fom4) = c("aln","month","year","season","vitD","clinic")

long_data = rbind(fom1, fom2, fom3, fom4)
```


## S.Burgess VitD Seasonal Correction BUT carried out by clinic

### FOM1

```{r}
## FOM1
temp_FOM1 = long_data %>% filter(clinic == "FOM1")
## Fit the default model
# fit = lm(vitD ~ season, data = temp_FOM1 )
fit = lm(vitD ~ as.factor(year) * season, data = temp_FOM1 )
## Fitted values
fit_ests = fitted(fit)
m = match(rownames(temp_FOM1), names(fit_ests))
## Seasonal Correction
# autumn_avg = predict(fit, newdata = list( season = "autumn") )
autumn_avg = predict(fit, newdata = list(year = 2009, season = "autumn") )
## Normalized vitD Estimates
temp_FOM1$vitD_normalized = (temp_FOM1$vitD - fit_ests[m]) + c(autumn_avg)

```


### FOM2

```{r}
## FOM2
temp_FOM2 = long_data %>% filter(clinic == "FOM2")
## Fit the default model
# fit = lm(vitD ~  season, data = temp_FOM2 )
fit = lm(vitD ~ as.factor(year) * season, data = temp_FOM2 )
## Fitted values
fit_ests = fitted(fit)
m = match(rownames(temp_FOM2), names(fit_ests))
## Seasonal Correction
# autumn_avg = predict(fit, newdata = list(season = "autumn") )
autumn_avg = predict(fit, newdata = list( year = 2012, season = "autumn") )
## Normalized vitD Estimates
temp_FOM2$vitD_normalized = (temp_FOM2$vitD - fit_ests[m]) + c(autumn_avg)
```


### FOM3

```{r}
## FOM3
temp_FOM3 = long_data %>% filter(clinic == "FOM3")
## Fit the default model
fit = lm(vitD ~ season, data = temp_FOM3 )
## Fitted values
fit_ests = fitted(fit)
m = match(rownames(temp_FOM3), names(fit_ests))
## Seasonal Correction
autumn_avg = predict(fit, newdata = list(season = "autumn") )
## Normalized vitD Estimates
temp_FOM3$vitD_normalized = (temp_FOM3$vitD - fit_ests[m]) + c(autumn_avg)
```

### FOM4

```{r}
## FOM4
temp_FOM4 = long_data %>% filter(clinic == "FOM4")
## Fit the default model
fit = lm(vitD ~ season, data = temp_FOM4 )
## Fitted values
fit_ests = fitted(fit)
m = match(rownames(temp_FOM4), names(fit_ests))
## Seasonal Correction
autumn_avg = predict(fit, newdata = list(season = "autumn") )
## Normalized vitD Estimates
temp_FOM4$vitD_normalized = (temp_FOM4$vitD - fit_ests[m]) + c(autumn_avg)
```


```{r}
long_data_v2 = rbind(temp_FOM1, temp_FOM2, temp_FOM3, temp_FOM4)
long_data$vitD_normalized = long_data_v2$vitD_normalized
```


## Normalize by clinic mean

```{r}
## Fit the  model
fit = lm(vitD_normalized ~ as.factor(clinic), data = long_data )

## Fitted values
fit_ests = fitted(fit)
m = match(rownames(long_data), names(fit_ests))

## Seasonal Correction
clinic_avg = predict(fit, newdata = list(clinic = "FOM1") )

## Normalized vitD Estimates
long_data$vitD_normalized_clinic_centered = (long_data$vitD_normalized - fit_ests[m]) + c(clinic_avg)
```




```{r}
P1 = long_data %>% ggplot(aes(x = month, y = vitD_normalized_clinic_centered)) +
  geom_smooth(method = gam, formula = y~s(x), aes(color = clinic) ) +
  facet_wrap(~clinic) +
  theme_bw()
```

```{r, fig.width = 6, fig.height = 5}
P2 = long_data %>% ggplot(aes(x = season, y = vitD_normalized_clinic_centered)) +
  geom_boxplot(aes(fill = as.factor(year)) ) +
  scale_fill_brewer(palette = "Set1") +
  facet_wrap(~ clinic) +
  theme_bw() + labs(fill = "year")
```


```{r}
corplot = function(wdata, clinicA, clinicB, PlotVar = "vitD_shifted" ){
  pcol = RColorBrewer::brewer.pal(8, "Set1")
  
  Adata = wdata %>% filter(clinic == clinicA )
  Bdata = wdata %>% filter(clinic == clinicB )
  
  ## match alns
  ids = sort( unique( c(Adata$aln, Bdata$aln) )  )
  
  m = match(ids, Adata$aln)
  Adata = Adata[m,]
  colnames(Adata) = paste0(colnames(Adata), "_", clinicA)
  
  m = match(ids, Bdata$aln)
  Bdata = Bdata[m,]
  colnames(Bdata) = paste0(colnames(Bdata), "_", clinicB)
  
  new_data = cbind(Adata, Bdata)
  
  x = paste0(PlotVar, "_", clinicA )
  y = paste0(PlotVar, "_", clinicB )
  ## correlation
  a = cor.test( new_data[, x], new_data[, y] )
  r = round( a$est, d =  4 )
  ### plot
  plot = new_data %>% ggplot(aes_string(x = x, y = y)) +
    geom_point(shape = 21, fill = pcol[2], size = 1.5, alpha = 0.5) + 
    geom_abline(intercept = 0, slope = 1, color = "black") +
    geom_smooth( method = lm, formula = y~x, color = pcol[1]) +
    # geom_smooth( method = gam, formula = y~s(x), color = pcol[3]) +
    labs(title = paste0("Pearson's r = ", r) ) +
    theme_bw()
  
  return(plot)
  
}
```

```{r, fig.width = 8, fig.height = 5}
p1 = corplot(wdata = long_data, clinicA = "FOM1", clinicB = "FOM2", PlotVar = "vitD_normalized_clinic_centered" )
p2 = corplot(wdata = long_data, clinicA = "FOM1", clinicB = "FOM3", PlotVar = "vitD_normalized_clinic_centered" )
p3 = corplot(wdata = long_data, clinicA = "FOM1", clinicB = "FOM4", PlotVar = "vitD_normalized_clinic_centered" )
p4 = corplot(wdata = long_data, clinicA = "FOM2", clinicB = "FOM3", PlotVar = "vitD_normalized_clinic_centered" )
p5 = corplot(wdata = long_data, clinicA = "FOM2", clinicB = "FOM4", PlotVar = "vitD_normalized_clinic_centered" )
p6 = corplot(wdata = long_data, clinicA = "FOM3", clinicB = "FOM4", PlotVar = "vitD_normalized_clinic_centered" )

PLOT = (p1 | p2 | p3) / (p4 | p5 | p6)
```


```{r, fig.width = 12, fig.height = 10, warning=FALSE}
plotout = (P1|P2)/PLOT + plot_layout(heights = c(1, 2))
plotout
```

## Write long data to file

```{r}
f = paste0(datadir, "working_data/vitD_long_data.txt")
write.table(long_data, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```

## Add the vitD_normalized_clinic_centered variable to the wide formatted data set

```{r}
## FOM1
temp = long_data %>% filter(clinic == "FOM1")
m = match(mydata$aln, temp$aln)
mydata$FOM1_vitD_season_and_clinic_centered = temp$vitD_normalized_clinic_centered[m]
```

```{r}
## FOM2
temp = long_data %>% filter(clinic == "FOM2")
m = match(mydata$aln, temp$aln)
mydata$FOM2_vitD_season_and_clinic_centered = temp$vitD_normalized_clinic_centered[m]
```

```{r}
## FOM3
temp = long_data %>% filter(clinic == "FOM3")
m = match(mydata$aln, temp$aln)
mydata$FOM3_vitD_season_and_clinic_centered = temp$vitD_normalized_clinic_centered[m]
```

```{r}
## FOM4
temp = long_data %>% filter(clinic == "FOM4")
m = match(mydata$aln, temp$aln)
mydata$FOM4_vitD_season_and_clinic_centered = temp$vitD_normalized_clinic_centered[m]
```

## Double Check

```{r}
cor.test(mydata$FOM1_vitD_season_and_clinic_centered, mydata$FOM2_vitD_season_and_clinic_centered )$estimate
```

## Write wide data to file

```{r}
f = paste0(datadir, "working_data/working_data_set_v1.txt")
write.table(mydata, file = f, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE)
```


